An Experimental Analysis of the Influence of the Precision of Parameters in an Artificial Neural Network

### 2025-11-18 paper/main.tex
weiter

Prior research in this domain can be broadly classified into two categories:

\subsubsection*{Quantization During Training}
Several studies have explored training neural networks with reduced precision
directly. Notable contributions include:
\begin{itemize}
	\item \textbf{Mixed Precision Training}, which leverages FP16 arithmetic for speed while maintaining FP32 master weights~\cite{micikevicius2017mixed}.
	\item \textbf{DoReFa-Net}, which quantizes weights, activations, and gradients to arbitrary bitwidths~\cite{zhou2018dorefa}.
	\item \textbf{Q-BERT}, a quantized BERT model that maintains accuracy using only 4-bit weights~\cite{shen2020qbert}.
\end{itemize}

\subsubsection*{Post-Training Quantization}
Other approaches focus on converting pretrained models into lower-precision
equivalents:
\begin{itemize}
	\item \textbf{Integer-Only Quantization} using symmetric quantization scales~\cite{jacob2018quantization}.
	\item \textbf{GPTQ}, a method for quantizing large transformer models without retraining~\cite{frantar2022gptq}.
	\item \textbf{SmoothQuant}, which scales activations to better match weight ranges during quantization~\cite{xiao2024smoothquant}.
\end{itemize}

\section{Artificial Neural Networks}
A typical neuron in a neural network performs a weighted sum of its inputs,
followed by an activation function which results in the neuron output $h$
depicted in~\eqref{eq:activation}. Here, $\omega\in\mathbb{R}^n$ is the weight
vector, $b \in \mathbb{R}$ is the bias, $x \in \mathbb{R}^n$ is the input, and
$\sigma$ is a nonlinear activation function (e.g., sigmoid or ReLU).

@@

\section{Memory Efficiency and Numerical Formats}
To reduce memory usage and computational load, precision-aware techniques such
as quantization and mixed-precision training are increasingly employed. Instead
of using 32-bit or 64-bit floating-point formats, parameters can be stored in
16-bit or even 8-bit fixed-point representations.
Equation~\ref{eq:floatingPointFormat} depicts the general floating point format
and terminology of significand, base, and exponent.

@@

\noindent
Tables~\ref{tab:binary-bits} and~\ref{tab:binary-precision} provide an overview
of the different floating point types and their respective characteristics. The
\emph{sign}, \emph{exponent}, and \emph{significand} fields specify how many
bits are allocated to encode the sign of the number, its scale, and its
significant digits, respectively. The \emph{exponent bias} is a constant added
to the true exponent so that it can be stored as an unsigned integer. The
\emph{precision} (in bits) and corresponding \emph{decimal digits} tell how
many significant digits can be represented accurately in each format.

@@

\begin{table}[H]
	\centering
	\begin{tabular}{@{}lccc@{}}
		\toprule
		\textbf{Type} & Exponent bias & Precision (bits) & Decimal digits \\
		\midrule
		Half (FP16)   & 15            & 11               & $\sim$3.3      \\
		Single (FP32) & 127           & 24               & $\sim$7.2      \\
		Double (FP64) & 1023          & 53               & $\sim$15.9     \\
		\bottomrule
	\end{tabular}
	\caption{Range and precision characteristics of the same formats (adapted from~\cite{fparithmetic}).}\label{tab:binary-precision}
\end{table}

@@

\section{Summary}
% \begin{itemize}
% 	\item \textbf{16-bit is enough for this classification task}: End-to-end FP16 preserved the accuracy.
% 	\item \textbf{Three decimals suffice}: Storing weights down to fractional length $3$ caused no visible accuracy loss.
% \end{itemize}

% \section{Discussion}
% The results of the first experiments have shown that either the selected task (MNIST
% classification) was not complex enough to be affected by the decrease in accuracy, or the
% accuracy would need to be decreased even further than FP16. Otherwise, the results
% generally indicate that networks trained and evaluated with as low as 16-bit precision can
% achieve comparable performance to their high-precision counterparts. Moreover, rounding
% pretrained weights to just 3 decimal digits preserves full accuracy, opening the door to
% aggressive post-training quantization.This is particularly relevant for edge devices and
% real-time applications where memory and compute constraints are more relevant.

>>>>

Prior research in this domain can be broadly divided into methods that use
quantization already during training and methods that quantize a pretrained
model later at deployment time.

\subsubsection*{Quantization During Training}
Mixed-precision training is an approach where most operations are performed in
FP16, but a separate high-precision copy of the parameters (often called
\textit{master weights}) is maintained in FP32~\cite{micikevicius2017mixed}.
The forward and backward passes use FP16 tensors for faster computation and
reduced memory usage, but parameter updates are applied to the FP32 master
weights. The updated FP32 weights are then cast back to FP16 for the next
forward pass. Keeping master weights in higher precision keeps the loss of
information smaller and helps the training stability.\\

\noindent
DoReFa-Net extends this concept by quantizing \textit{weights},
\textit{activations}, and \textit{gradients} to low-precision
formats~\cite{zhou2018dorefa}. The method allows using \textit{arbitrary
	bitwidths}, i.e., the number of bits is not restricted to standard IEEE
floating-point formats, but can be any small integer (for example, 1, 2, or 4
bits for weights and slightly higher bitwidths for activations and gradients).
This is flexible and makes it possible to set the precision specific to
hardware constraints or application requirements, but the training of the
network is still end-to-end.\\

\noindent
Another work is Q-BERT~\cite{shen2020qbert}, that focuses on transformer-based
language models. BERT (Bidirectional Encoder Representations from Transformers)
is a large transformer architecture pretrained on large text data and used as a
generic language representation model. Q-BERT shows that the weights of BERT
can be quantized down to 4 bits while maintaining accuracy close to the
full-precision model.

\subsubsection*{Post-Training Quantization}
Post-training quantization begins with an already trained full-precision model
and converts it to a lower precision without (or with minimal) additional
training. An often used approach is integer-only quantization with symmetric
scaling~\cite{jacob2018quantization}. Here, both weights and activations are
mapped to signed integers (for example 8-bit) using a scale factor that is the
same across positive and negative values (therefore symmetric). This design
allows the entire inference flow to be implemented with integer arithmetic,
which is better for embedded or mobile hardware.\\

\noindent
Methods such as GPTQ (Gradient Post-Training Quantization) have been tested to
quantize large transformer models. Examples are GPT language models with
minimal or no retraining~\cite{frantar2022gptq}. GPTQ treats quantization as an
optimization problem by using approximations of the local curvature to choose
quantization levels that minimize the error introduced by replacing
full-precision weights with low-precision ones. That makes this approach better
for very large models where a full retraining with quantization would be too
expensive.\\

\noindent
SmoothQuant focuses more on activations~\cite{xiao2024smoothquant}. It rescales
activations and compensates for this rescaling in the weight matrices so that
the activation distributions become more smooth and better aligned with the
range supported by low-precision integer formats. This improves the robustness
of post-training quantization, especially in transformer architectures where
activation outliers can otherwise complicate the scaling and lead to large
quantization errors.

\subsection{Artificial Neural Networks}
A typical neuron in a neural network performs a weighted sum of its inputs,
followed by an activation function which results in the neuron output $h$
depicted in~\eqref{eq:activation}. Here, $\omega\in\mathbb{R}^n$ is the weight
vector, $b \in \mathbb{R}$ is the bias, $x \in \mathbb{R}^n$ is the input, and
$\sigma$ is a nonlinear activation function (e.g., sigmoid or ReLU).

@@

\subsection{Memory Efficiency and Numerical Formats}
To reduce memory usage and computational load, precision-aware techniques such
as quantization and mixed-precision training are increasingly employed. Instead
of using 32-bit or 64-bit floating-point formats, parameters can be stored in
16-bit or even 8-bit fixed-point representations.
Equation~\ref{eq:floatingPointFormat} depicts the general floating point format
and terminology of significand, base, and exponent.

@@

\noindent
Tables~\ref{tab:binary-bits} and~\ref{tab:binary-precision} provide an overview
of the different floating point types and their respective characteristics. The
\textit{sign}, \textit{exponent}, and \textit{significand} fields specify how many
bits are allocated to encode the sign of the number, its scale, and its
significant digits, respectively. The \textit{exponent bias} is a constant added
to the true exponent so that it can be stored as an unsigned integer. The
\textit{precision} (in bits) and corresponding \textit{decimal digits} tell how
many significant digits can be represented accurately in each format.

@@

\begin{table}[H]
	\centering
	\begin{tabular}{@{}lccc@{}}
		\toprule
		\textbf{Type} & Exponent bias & Precision (bits) & Decimal digits \\
		\midrule
		Half (FP16)   & 15            & 11               & $\sim$3.3      \\
		Single (FP32) & 127           & 24               & $\sim$7.2      \\
		Double (FP64) & 1023          & 53               & $\sim$15.9     \\
		\bottomrule
	\end{tabular}
	\caption{Range and precision characteristics of the formats (adapted from~\cite{fparithmetic}).}\label{tab:binary-precision}
\end{table}

@@

\section{Summary}
